---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r global_options, include=FALSE, echo=FALSE,message=FALSE,results="hide"}
# set the figures directory to correspond to the Course instructions
knitr::opts_chunk$set(fig.path='figure/')
```

## Loading and preprocessing the data
```{r activity_monitoring}
ACT <- read.csv(unz("activity.zip","activity.csv"),na.strings="NA",colClasses=c("numeric","Date","numeric"))
```
The variables included in this dataset (variable "ACT") are:

1. **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)
2. **date**: The date on which the measurement was taken in YYYY-MM-DD format. Gets converted to an R Date class value.
3. **interval**: Identifier for the 5-minute interval in which measurement was taken


## What is mean total number of steps taken per day?
```{r histogram_prep,echo=FALSE,results="hide",message=FALSE}
library(dplyr)
```
Compute the total number of steps per day and store in the R object "ACT_s".
```{r summarize_data}
ACT_s <- ACT %>% group_by(date) %>% summarise(steps.day=sum(steps,na.rm=TRUE))
```
Following is a chart of the number of days of different reported activity levels
```{r histogram}
hist(ACT_s$steps.day,breaks=10,col="wheat",main="Total steps per day",
      xlab="Total steps per day",ylab="Number of days with total steps")
```      

```{r mean and median, results="as is"}
ACT_s_mean <- mean(ACT_s$steps.day)
ACT_s_median <- median(ACT_s$steps.day)
```      

The mean total number of steps per day is `r sprintf("%5.0f",ACT_s_mean)` (whereas the median is `r sprintf("%5.0f",ACT_s_median)`).      

## What is the average daily activity pattern?
Compute the average number of steps per interval across all days and store in the R object "ACT_a".
```{r avg_number_of_steps_interval}
ACT_a <- ACT %>% group_by(interval) %>% summarise(steps.interval=mean(steps,na.rm=TRUE))
```
Create a new variable in "ACT_a" called "time_of_interval" of class "hms" to help ggplot to understand how to plot the time-series.
```{r avg_nr_steps_prep,echo=FALSE,results="hide",message=FALSE}
library(hms)
library(ggplot2)
```
```{r convert_interval_to_time}
ACT_a$time_of_interval <- hms(rep(0,length(ACT_a$interval)),ACT_a$interval%%100,floor(ACT_a$interval/100))
```
Here, then, a chart of the average number of steps per interval
```{r time_series}
ggplot(ACT_a,aes(x=time_of_interval,y=steps.interval)) +
    geom_line() +
    scale_x_time(breaks=seq(0,24*3600,4*3600))+
    xlab("Time of day") +
    ylab("Average steps in 5-min interval") + 
    ggtitle("Activity at different times of the day")
```

In this chart you can see that the time interval with the largest average number of steps over the days is:

```{r max_steps}
idx_max_steps <- which.max(ACT_a$steps.interval)
time_max_steps <- ACT_a$time_of_interval[idx_max_steps]
```

The largest average number of steps across the days happens in the 5-min interval that starts at `r time_max_steps` hours.

## Imputing missing values

The dataset has missing values. Below we investigate how many.

```{r missing_steps,results="hide"}
missing_steps <- is.na(ACT$steps)
```

The number of missing observations of steps is `r sum(missing_steps)`.

The next code sections will replace the missing step counts with the average number of steps for the corresponding interval across all days measured. The R object "ACT_filled" will contain the adjusted dataset. The new, adjusted, dataset will be summarized as the R object "ACT_filled_s"

```{r missing_steps_fill}
ACT_filled <- ACT
for (i in 1:length(ACT$steps)) {
    if (is.na(ACT_filled$steps[i])) {
        ACT_filled$steps[i] <- ACT_a[ACT_a$interval==ACT_filled$interval[i],]$steps.interval
    }
}
```
```{r summarize_data_filled}
ACT_filled_s <- ACT_filled %>% group_by(date) %>% summarise(steps.day=sum(steps,na.rm=TRUE))
```

Following is a chart of the number of days of different reported activity levels:

```{r histogram_filled}
hist(ACT_filled_s$steps.day,breaks=10,col="orange",main="Total steps per day (after NA replacement)",
      xlab="Total steps per day",ylab="Number of days with total steps")
```      

```{r mean and median_filled}
ACT_filled_mean_steps <- mean(ACT_filled_s$steps.day)
ACT_filled_median_steps <- median(ACT_filled_s$steps.day)
```

The adjusted dataset's mean (`r sprintf("%5.0f",ACT_filled_mean_steps)`) and median 
(`r sprintf("%5.0f",ACT_filled_median_steps)`) total steps per day are now the same and larger than when the NA were still present.

## Are there differences in activity patterns between weekdays and weekends?

The following code introduces an additional variable "weekday" to capture whether the observations are for a Mon-Fri "weekday" or a Sat-Sun "weekend" day. It also summarizes the average steps per day into the R object "ACT_filled_a" by the category "weekday" just described.

```{r facet_prep}
ACT_filled$weekday <- ifelse(weekdays(ACT_filled$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
ACT_filled_a <- ACT_filled %>% group_by(weekday,interval) %>% summarise(steps.interval=mean(steps))
ACT_filled_a$time_of_interval <- hms(rep(0,length(ACT_filled_a$interval)),
                                     ACT_filled_a$interval%%100,floor(ACT_filled_a$interval/100))
```

With the summarized data its possible to observe some differences between activity on a "weekday" and a "weekend" day.

```{r facet_plot}
ggplot(ACT_filled_a,aes(x=time_of_interval,y=steps.interval)) +
    geom_line() +
    scale_x_time(breaks=seq(0,24*3600,4*3600))+
    xlab("Time of day") +
    ylab("Average steps in 5-min interval") + 
    ggtitle("Activity at different times of the day") +
    facet_grid(weekday ~ .)
```

